---
title: "Analyzing-WASH-data"
output: html_document
---

Load packages and the data
```{R}
library(devtools)
load_all('/home/salbalkus/Research/RieszCML/')

library(nadir)
library(readr)
library(tidyverse)
library(glmnet)
library(here)
# library(vaccine)


```

First, load and clean the vaccine trial data.
```{R}

large_data <- read_csv("https://raw.githubusercontent.com/nhejazi/pub_txshift_biometrics/refs/heads/master/application/data/primary505_for_sharing.csv") %>%
  select(pub_id, HIVwk28preunbl, trt, age, racecc, BMI, bhvrisk)

subsample_data<- read_csv("https://raw.githubusercontent.com/nhejazi/pub_txshift_biometrics/refs/heads/master/application/data/janes_ipc_weights.csv") %>%
  select(pub_id,                              # for censoring indicator
         cd4_score_scaled, cd8_score_scaled,  # immune markers from Janes et al
         wt, wt_hal,                          # IPC weights (true and estimated)
        )

#nuisances <- readRDS(here("vignettes", "cd4_nuisance_delta_grid.rds"))


df <- left_join(large_data, subsample_data, by = "pub_id") %>% filter(trt == 1)
df$sampled <- is.na(df$cd4_score_scaled) == 0
df$wt[df$sampled == 0] <- 0
df$wt_hal[df$sampled == 0] <- 0
df$Y <- df$HIVwk28preunbl

df_ct <- df %>% drop_na()

form_m <- Y ~ cd4_score_scaled + age + racecc + BMI + bhvrisk

m_fit <- super_learner(
      data = df_ct,
      formulas = form_m,
      outcome_type = "binary",
      n_folds = 10,
      learners = list(
        lnr_gam, lnr_hal,
        lnr_ranger, lnr_ranger, lnr_ranger, lnr_ranger 
        ),
      extra_learner_args = list(
        NULL, list(smoothness_orders = 0, num_knots = 50, max_degree = 2, reduce_basis = 0.08),
        NULL, list(alpha = 0.05), list(max.depth = 6), list(max.depth = 3)
      )
    )

```

Next, we will Riesz represent the parameter of interest, which is the mean outcome under a shift in the treatment variable, under complete data.

```{R}
rc <- function(delta, name){return(RieszCurve$new(
  nuis = function(df) {

    df_ct <- df %>% drop_na()

    # regression for Y given current history
    form_m <- Y ~ get(name) + age + racecc + BMI + bhvrisk
    #m_fit <- glm(form_m,
    #             family = binomial(),
    #             data = df_ct)
    m_fit <- super_learner(
      data = df_ct,
      formulas = form_m,
      outcome_type = "binary",
      n_folds = 10,
      learners = list(
        lnr_gam, lnr_hal,
        lnr_ranger, lnr_ranger, lnr_ranger, lnr_ranger 
        ),
      extra_learner_args = list(
        NULL, list(smoothness_orders = 0, num_knots = 50, max_degree = 2, reduce_basis = 0.05),
        NULL, list(alpha = 0.05), list(max.depth = 6), list(max.depth = 3)
      )
    )

    # treatment mechanism for A2
    df_inv_shift <- df_shift <- df_ct
    df_shift[[name]] = df_shift[[name]] + delta
    df_inv_shift[[name]] = df_shift[[name]] - delta
    df_aug <- rbind(cbind(df_ct, I = 0), cbind(df_inv_shift, I = 1))
    
    form_r = I ~ get(name) + age + racecc + BMI + bhvrisk
    #r_fit <- glm(form_r,
    #             family = binomial(),
    #             data = df_aug)
    r_fit <- super_learner(
      data = df_aug,
      formulas = form_r,
      outcome_type = "binary",
      n_folds = 10,
      learners = list(
        lnr_gam, lnr_hal,
        lnr_ranger, lnr_ranger, lnr_ranger, lnr_ranger 
        ),
      extra_learner_args = list(
        NULL, list(smoothness_orders = 0, num_knots = 50, max_degree = 2, reduce_basis = 0.05),
        NULL, list(alpha = 0.05), list(max.depth = 6), list(max.depth = 3)
      )
    )

    list(
      m = predict(m_fit, newdata = df_ct, type = "response"),
      m1 = predict(m_fit, newdata = df_shift, type = "response"),
      r = predict(r_fit, newdata = df_inv_shift, type = "response")
    )
  },
  h = ~ m1,
  f = ~ m,
  alpha = ~ r / (1 - r),
  ic_expr = ~ h + alpha * (Y - f),
  targeting_steps = list(
    m = list(),
    m1 = list(set = setNames(list(df_ct[[name]] + delta), name))
  )
))}

est_tmle_complete <- riesz_tmle(
  data = df_ct,
  rc = rc(1, "cd4_score_scaled"),
  fluctuation_type = "logistic",
  bounds = c(min(df_ct$Y), max(df_ct$Y)),
  outcome_col = "Y"
)

```

Then, we will use the complete data EIF estimate to perform a TMLE for the two-phase sampling data.
```{r}

two_phase_tmle <- function(df, df_ct, delta, name){

  # First, estimate the TMLE in the complete data
  est_tmle_complete <- riesz_tmle(
    data = df_ct,
    rc = rc(delta, name),
    fluctuation_type = "logistic",
    bounds = c(min(df_ct$Y), max(df_ct$Y)),
    outcome_col = "Y"
  )

  # Add the estimated EIF from the previous step to the data
  df_updated <- left_join(df %>% select(-Y), data.frame(pub_id = df_ct$pub_id, Y = est_tmle_complete$h_star), by = "pub_id")
  # Replace NAs so they don't break anything when we multiply by an indicator
  df_updated["casecontrol"] <- ifelse(is.na(df_updated[name]), 0, 1)
  df_updated$Y[is.na(df_updated$Y)] <- 0
  df_updated$wt[is.na(df_updated$wt)] <- 0

  # Implement Riesz representation for the two-phase sampling data, using the EIF from the complete 
  # data as the "outcome" in the second stage TMLE
  rc_cc <- RieszCurve$new(
    nuis = function(df_updated) {
      # regression for Y given current history
      m_fit <- glm(Y ~ HIVwk28preunbl + casecontrol,
                  family = gaussian(),
                  data = df_updated)
      # we already have IC weights, so we can just use those as the "propensity score" for the shift
      # m and m1 are the same since we only operate in the complete data
      df_updated_shift = df_updated
      df_updated_shift["casecontrol"] <- 1
      list(
        m = predict(m_fit, newdata = df_updated, type = "response"),
        m1 = predict(m_fit, newdata = df_updated, type = "response"),
        invprop = df_updated$wt
      )
    },
    h = ~ m1,
    f = ~ m,
    alpha = ~ invprop,
    ic_expr = ~ h + alpha * (Y - f),
    targeting_steps = list(
      m = list(),
      m1 = list(set = list(casecontrol = 1))
    )
  )

  est_tmle <- riesz_tmle(
    data = df_updated,
    rc = rc_cc,
    fluctuation_type = "logistic",
    bounds = c(min(df_updated$Y), max(df_updated$Y)),
    outcome_col = "Y"
  )

  return(est_tmle)
}

#two_phase_tmle(df, df_ct, 0, "cd8_score_scaled")
```

We can now run this estimator over various shifts in the data

```{r}
set.seed(2)
shifts <- c(-2, -1.5, -1, -0.5, 0, 0.5, 1, 1.5, 2)
cd4_results <- lapply(shifts, \(x) two_phase_tmle(df, df_ct, x, "cd4_score_scaled"))
cd4_ests <- sapply(cd4_results, \(x) c(est = x$estimate, lo = x$ci_low, hi = x$ci_high))

cd4_df <- data.frame(
  shift = shifts,
  est = cd4_ests[1, ],
  lo = cd4_ests[2, ],
  hi = cd4_ests[3, ]
)

ggplot(cd4_df, aes(x = shift, y = est)) +
  geom_point() +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.1) +
  labs(
    title = "CD4 Estimates by Shift",
    x = "Shift",
    y = "Estimate"
  ) +
  theme_minimal()

ggsave("plot.png")
```


Or, we could follow the Rose and van der Laan strategy of weighting the TMLE update.

```
