---
title: "ATE and ATT with RieszCML"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ATE and ATT with RieszCML}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette illustrates estimation of the average treatment effect (ATE) and average treatment effect among the treated (ATT) using `{RieszCML}`.

We simulate data from a simple point-treatment causal model where the truth is known:
$$
L \sim N(0,1), \qquad
A \mid L \sim \text{Bernoulli}\{\operatorname{expit}(L)\},$$
$$Y = 2 + L + \tau A + \varepsilon, \qquad \varepsilon \sim N(0, \sigma^2).$$
Because the treatment effect is additive and constant, the causal contrasts are known exactly:
$$\mathbb E[Y^1] = 2 + \mathbb E[L] + \tau = 2 + \tau,$$
$$\mathbb E[Y^0] = 2 + \mathbb E[L] = 2,$$
$$\text{ATE} = \mathbb E[Y^1 - Y^0] = \tau,
\qquad
\text{ATT} = \mathbb E[Y^1 - Y^0 \mid A=1] = \tau.$$
```{r}
library(RieszCML)

set.seed(1)

n <- 5000
tau <- 1.5
sigma <- 0.25

L <- rnorm(n)
A <- rbinom(n, size = 1, prob = plogis(L))
Y <- 2 + L + tau * A + rnorm(n, sd = sigma)

df <- data.frame(L = L, A = A, Y = Y)

truth_mu1 <- 2 + tau
truth_mu0 <- 2
truth_ate <- tau
truth_att <- tau

c(
  truth_mu1 = truth_mu1,
  truth_mu0 = truth_mu0,
  truth_ate = truth_ate,
  truth_att = truth_att
)
```

We first estimate $\mathbb E[Y^1]$ and $\mathbb E[Y^0]$, then combine them via the delta method.

```{r}
rc_mu1 <- riesz_curve_catalog$cfmean_a(
  a = 1,
  nuis = function(data) {
    m_fit <- lm(Y ~ L + A, data = data)
    g_fit <- glm(I(A == 1) ~ L, family = binomial(), data = data)

    data1 <- data
    data1$A <- 1

    list(
      m = predict(m_fit, newdata = data),
      ma = predict(m_fit, newdata = data1),
      g = predict(g_fit, newdata = data, type = "response")
    )
  }
)

rc_mu0 <- riesz_curve_catalog$cfmean_a(
  a = 0,
  nuis = function(data) {
    m_fit <- lm(Y ~ L + A, data = data)
    g_fit <- glm(I(A == 0) ~ L, family = binomial(), data = data)

    data0 <- data
    data0$A <- 0

    list(
      m = predict(m_fit, newdata = data),
      ma = predict(m_fit, newdata = data0),
      g = predict(g_fit, newdata = data, type = "response")
    )
  }
)

fit_mu1 <- riesz_estimate(df, rc_mu1)
fit_mu0 <- riesz_estimate(df, rc_mu0)

fit_mu1
fit_mu0

fit_ate <- riesz_delta_difference(
  fit_mu1,
  fit_mu0,
  parameter = "ATE = E[Y^1] - E[Y^0]"
)

fit_ate

c(
  estimate_mu1 = fit_mu1$estimate,
  truth_mu1 = truth_mu1,
  estimate_mu0 = fit_mu0$estimate,
  truth_mu0 = truth_mu0,
  estimate_ate = fit_ate$estimate,
  truth_ate = truth_ate
)
```

Of course, that would be equivalent to using the `ate` method in the `riesz_curve_catalog`. 

```{r}
rc_ate <- riesz_curve_catalog$ate(
  nuis = function(data) {
    m_fit <- lm(Y ~ L + A, data = data)
    g_fit <- glm(A ~ L, family = binomial(), data = data)

    data1 <- data
    data0 <- data
    data1$A <- 1
    data0$A <- 0

    list(
      m = predict(m_fit, newdata = data),
      m1 = predict(m_fit, newdata = data1),
      m0 = predict(m_fit, newdata = data0),
      g = predict(g_fit, newdata = data, type = "response")
    )
  }
)

fit_ate <- riesz_estimate(df, rc_ate)
fit_ate

c(
  estimate_ate = fit_ate$estimate,
  truth_ate = truth_ate
)
```

```{r}
rc_att <- riesz_curve_catalog$att(
  nuis = function(data) {
    m_fit <- lm(Y ~ L + A, data = data)
    g_fit <- glm(A ~ L, family = binomial(), data = data)

    data1 <- data
    data0 <- data
    data1$A <- 1
    data0$A <- 0

    list(
      m = predict(m_fit, newdata = data),
      m1 = predict(m_fit, newdata = data1),
      m0 = predict(m_fit, newdata = data0),
      g = predict(g_fit, newdata = data, type = "response")
    )
  }
)

fit_att <- riesz_estimate(df, rc_att)
fit_att

c(
  estimate_att = fit_att$estimate,
  truth_att = truth_att
)
```

```{r}

```
