---
title: "Composed-Riesz-Missingness-and-Longitudinal"
output: html_document
---

One of the main motivations for the RieszCML framework is that many interesting estimands can be built compositionally from simpler ones.

This vignette illustrates that idea with three examples:

  - ATE with missing outcome under MAR as a two-stage procedure,
  - a two-timepoint longitudinal regime mean as a composed object,
  - a longitudinal regime contrast with missing final outcome as a three-stage composed object plus delta method.

```{r}
library(RieszCML)
```

## ATE under MAR 

```{r}

set.seed(11)

n <- 5000
tau <- 1.5
sigma <- 0.25

L <- rnorm(n)
A <- rbinom(n, 1, plogis(L))
Y_full <- 2 + L + tau * A + rnorm(n, sd = sigma)
R <- rbinom(n, 1, plogis(-0.5 + 0.5 * L + 0.3 * A))

Y <- ifelse(R == 1, Y_full, 0)

df <- data.frame(L = L, A = A, R = R, Y = Y)

truth_mu1 <- 2 + tau
truth_mu0 <- 2
truth_ate <- tau

c(truth_mu1 = truth_mu1, truth_mu0 = truth_mu0, truth_ate = truth_ate)
```

```{r}
rc_missing <- riesz_curve_catalog$missing_mean(
  L = "L",
  R = "R",
  Y = "Y",
  nuis = function(data) {
    obs <- data[data$R == 1, , drop = FALSE]

    m_fit <- lm(Y ~ L + A, data = obs)
    pi_fit <- glm(R ~ L + A, family = binomial(), data = data)

    list(
      m = predict(m_fit, newdata = data),
      pi = predict(pi_fit, newdata = data, type = "response")
    )
  }
)

make_rc_treatment_mean <- function(a) {
  alpha_formula <- as.formula(paste0("~ I(A == ", a, ")/g"))

  RieszCurve$new(
    nuis = function(data) {
      m_fit <- lm(`h[j=1]` ~ L + A, data = data)
      g_fit <- glm(I(A == a) ~ L, family = binomial(), data = data)

      data_a <- data
      data_a$A <- a

      list(
        m = predict(m_fit, newdata = data),
        ma = predict(m_fit, newdata = data_a),
        g = predict(g_fit, newdata = data, type = "response")
      )
    },
    alpha = alpha_formula,
    f = ~ m,
    h = ~ ma,
    ic_expr = ~ h + alpha * (`h[j=1]` - f)
  )
}

rc_mu1_composed <- ComposedRieszCurve$new(
  rc_list = list(
    rc_missing,
    make_rc_treatment_mean(a = 1)
  )
)

rc_mu0_composed <- ComposedRieszCurve$new(
  rc_list = list(
    rc_missing,
    make_rc_treatment_mean(a = 0)
  )
)

fit_mu1 <- riesz_estimate(df, rc_mu1_composed)
fit_mu0 <- riesz_estimate(df, rc_mu0_composed)

fit_mu1
fit_mu0

fit_ate <- riesz_delta_difference(
  fit_mu1,
  fit_mu0,
  parameter = "ATE with MAR outcome missingness"
)

fit_ate

c(
  estimate_mu1 = fit_mu1$estimate,
  truth_mu1 = truth_mu1,
  estimate_mu0 = fit_mu0$estimate,
  truth_mu0 = truth_mu0,
  estimate_ate = fit_ate$estimate,
  truth_ate = truth_ate
)
```

## Example 2: Two-Timepoint Longitudinal Mean

```{r}
set.seed(12)

n <- 5000

beta0 <- 0.5
beta1 <- 0.7
beta2 <- 2.0

theta0 <- 1.0
theta2 <- 1.5
theta3 <- 3.0

L1 <- rnorm(n)
A1 <- rbinom(n, 1, plogis(-0.2 + 1.0 * L1))
U2 <- rnorm(n)
L2 <- beta0 + beta1 * L1 + beta2 * A1 + U2
A2 <- rbinom(n, 1, plogis(-0.3 + 0.6 * A1 + 1.2 * L2))
Y <- theta0 + theta2 * L2 + theta3 * A2 + rnorm(n)

df_long <- data.frame(L1 = L1, A1 = A1, L2 = L2, A2 = A2, Y = Y)

truth_mu11 <- theta0 + theta2 * (beta0 + beta2) + theta3
truth_mu11

rc_A2_1 <- RieszCurve$new(
  nuis = function(data) {
    m_fit <- lm(Y ~ L1 + A1 + L2 + A2, data = data)
    g_fit <- glm(A2 ~ L1 + A1 + L2, family = binomial(), data = data)

    data1 <- data
    data1$A2 <- 1

    list(
      m = predict(m_fit, newdata = data),
      m1 = predict(m_fit, newdata = data1),
      g = predict(g_fit, newdata = data, type = "response")
    )
  },
  h = ~ m1,
  f = ~ m,
  alpha = ~ A2 / g,
  ic_expr = ~ h + alpha * (Y - f)
)

rc_A1_1 <- RieszCurve$new(
  nuis = function(data) {
    m_fit <- lm(`h[j=1]` ~ L1 + A1, data = data)
    g_fit <- glm(A1 ~ L1, family = binomial(), data = data)

    data1 <- data
    data1$A1 <- 1

    list(
      m = predict(m_fit, newdata = data),
      m1 = predict(m_fit, newdata = data1),
      g = predict(g_fit, newdata = data, type = "response")
    )
  },
  h = ~ m1,
  f = ~ m,
  alpha = ~ A1 / g,
  ic_expr = ~ h + alpha * (`h[j=1]` - f)
)

rc_mu11 <- ComposedRieszCurve$new(
  rc_list = list(rc_A2_1, rc_A1_1)
)

fit_mu11 <- riesz_estimate(df_long, rc_mu11)
fit_mu11

c(
  estimate_mu11 = fit_mu11$estimate,
  truth_mu11 = truth_mu11
)
```

## Example 3: Longitudinal Regime with Missing Final Outcome


```{r}
set.seed(13)

n <- 5000

beta0 <- 0.5
beta1 <- 0.7
beta2 <- 2.0

theta0 <- 1.0
theta2 <- 1.5
theta3 <- 0.8
theta4 <- 2.2

L1 <- rnorm(n)
A1 <- rbinom(n, 1, plogis(-0.2 + 1.0 * L1))
U2 <- rnorm(n)
L2 <- beta0 + beta1 * L1 + beta2 * A1 + U2
A2 <- rbinom(n, 1, plogis(-0.3 + 0.6 * A1 + 1.2 * L2))

Y_full <- theta0 + theta2 * L2 + theta3 * A1 + theta4 * A2 + rnorm(n)
R <- rbinom(n, 1, plogis(-0.4 + 0.3 * L1 + 0.2 * A1 + 0.4 * L2 + 0.2 * A2))
Y <- ifelse(R == 1, Y_full, 0)

df_long_miss <- data.frame(L1 = L1, A1 = A1, L2 = L2, A2 = A2, R = R, Y = Y)

truth_mu11 <- theta0 + theta2 * (beta0 + beta2 * 1) + theta3 * 1 + theta4 * 1
truth_mu00 <- theta0 + theta2 * (beta0 + beta2 * 0) + theta3 * 0 + theta4 * 0
truth_contrast <- truth_mu11 - truth_mu00

c(
  truth_mu11 = truth_mu11,
  truth_mu00 = truth_mu00,
  truth_contrast = truth_contrast
)

rc_R <- riesz_curve_catalog$missing_mean(
  L = "L1",
  R = "R",
  Y = "Y",
  nuis = function(data) {
    obs <- data[data$R == 1, , drop = FALSE]

    m_fit <- lm(Y ~ L1 + A1 + L2 + A2, data = obs)
    pi_fit <- glm(R ~ L1 + A1 + L2 + A2, family = binomial(), data = data)

    list(
      m = predict(m_fit, newdata = data),
      pi = predict(pi_fit, newdata = data, type = "response")
    )
  }
)

make_rc_A2 <- function(a2) {
  alpha_formula <- as.formula(paste0("~ I(A2 == ", a2, ")/g"))

  RieszCurve$new(
    nuis = function(data) {
      m_fit <- lm(`h[j=1]` ~ L1 + A1 + L2 + A2, data = data)
      g_fit <- glm(I(A2 == a2) ~ L1 + A1 + L2, family = binomial(), data = data)

      data_a2 <- data
      data_a2$A2 <- a2

      list(
        m = predict(m_fit, newdata = data),
        ma = predict(m_fit, newdata = data_a2),
        g = predict(g_fit, newdata = data, type = "response")
      )
    },
    alpha = alpha_formula,
    f = ~ m,
    h = ~ ma,
    ic_expr = ~ h + alpha * (`h[j=1]` - f)
  )
}

make_rc_A1 <- function(a1) {
  alpha_formula <- as.formula(paste0("~ I(A1 == ", a1, ")/g"))

  RieszCurve$new(
    nuis = function(data) {
      m_fit <- lm(`h[j=2]` ~ L1 + A1, data = data)
      g_fit <- glm(I(A1 == a1) ~ L1, family = binomial(), data = data)

      data_a1 <- data
      data_a1$A1 <- a1

      list(
        m = predict(m_fit, newdata = data),
        ma = predict(m_fit, newdata = data_a1),
        g = predict(g_fit, newdata = data, type = "response")
      )
    },
    alpha = alpha_formula,
    f = ~ m,
    h = ~ ma,
    ic_expr = ~ h + alpha * (`h[j=2]` - f)
  )
}

rc_mu11 <- ComposedRieszCurve$new(
  rc_list = list(
    rc_R,
    make_rc_A2(a2 = 1),
    make_rc_A1(a1 = 1)
  )
)

rc_mu00 <- ComposedRieszCurve$new(
  rc_list = list(
    rc_R,
    make_rc_A2(a2 = 0),
    make_rc_A1(a1 = 0)
  )
)

fit_mu11 <- riesz_estimate(df_long_miss, rc_mu11)
fit_mu00 <- riesz_estimate(df_long_miss, rc_mu00)

fit_mu11
fit_mu00

fit_regime_contrast <- riesz_delta_difference(
  fit_mu11,
  fit_mu00,
  parameter = "E[Y^(1,1)] - E[Y^(0,0)] with MAR final outcome"
)

fit_regime_contrast

c(
  estimate_mu11 = fit_mu11$estimate,
  truth_mu11 = truth_mu11,
  estimate_mu00 = fit_mu00$estimate,
  truth_mu00 = truth_mu00,
  estimate_contrast = fit_regime_contrast$estimate,
  truth_contrast = truth_contrast
)
```
