---
title: "Getting Started with RieszCML: Riesz-Representer-Based Estimation from First Principles"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting Started with RieszCML}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(RieszCML)
```

`{RieszCML}` constructs asymptotically efficient, doubly robust estimators —
one-step estimators and targeted minimum loss-based estimators (TMLEs) — for
any statistical parameter that can be written as a **linear functional of one
or more nuisance functions**. This covers a remarkable share of the estimands
used in causal inference and missing-data problems: counterfactual means,
average treatment effects (ATE), effects among the treated (ATT), controlled
direct effects (CDE), modified treatment policies (MTPs), longitudinal
treatment regimes (including LMTPs), subgroup effects, average derivative
effects, and parameters corrected for outcome missingness or two-phase
sampling.

The package is in part a menu of estimands. However, it also gives you an
algebra of building blocks for constructing efficient estimators:

* `RieszCurve` — a declarative description of one linear functional: its
  nuisance functions, its Riesz representer, and its efficient influence
  function (EIF);
* `ComposedRieszCurve` — a chain of `RieszCurve`s for *nested* functionals
  (sequential regressions, as in longitudinal problems and mediation-style
  parameters);
* `riesz_estimate()` — the one-step (de-biased plug-in) estimator;
* `riesz_tmle()` — the TMLE, a substitution estimator that fluctuates the
  fitted regressions until the EIF estimating equation is solved;
* `riesz_delta_*()` — delta-method combinators for contrasts, ratios, and
  odds ratios of fitted linear functionals;
* `riesz_curve_catalog` — pre-specified curves for common estimands, which
  double as worked examples of the interface.

The goal of this vignette is that, by the end, users can Riesz-represent an
estimand they care about, as opposed to being restricted to what's in the catalog, 
and implement a correct one-step estimator and TMLE for it themselves. To get
there we first need about two pages of theory, because each argument of
`RieszCurve$new()` corresponds to an object in that theory.

## 1. The theory in two pages

### 1.1 Linear functionals and Riesz representers

Suppose we observe $n$ i.i.d. copies of $O \sim P_0$ and want to estimate a
parameter of the form

$$
\psi_0 \;=\; \Psi(\eta_0) \;=\; E_{P_0}\!\left[\, h(O;\, \eta_0) \,\right],
$$

where $\eta_0$ is a *nuisance function* (usually a regression, e.g.
$m_0(A, L) = E[Y \mid A, L]$) and $h$ is a fixed transformation of it that is
**linear** in $\eta$. The prototypical example is the counterfactual mean
under treatment,

$$
\psi_0 = E\big[\, m_0(1, L) \,\big],
\qquad
h(O; m) = m(1, L),
$$

where $h$ is "evaluate the regression at $A = 1$" — a linear operation.

Because $\Psi$ is a bounded linear functional of $\eta$, the Riesz
representation theorem guarantees a unique function $\alpha_0$ — the **Riesz
representer** — such that

$$
\Psi(\eta) = E_{P_0}\big[\, \alpha_0(X)\, \eta(X) \,\big]
\quad \text{for every } \eta,
$$

where $X$ denotes the arguments of $\eta$. The representer is a
change-of-measure object: whenever the parameter integrates $\eta$ against a
distribution $P^\star$ that one cannot sample from (a counterfactual world, the
full-data distribution behind a missingness process, a shifted exposure
distribution), $\alpha_0 = dP^\star / dP_0$ is the Radon–Nikodym derivative
that re-expresses that integral under the observed-data distribution. This is
why every representer looks like a weight:

| Parameter | $h(O; \eta)$ | Riesz representer $\alpha(O)$ |
|---|---|---|
| Counterfactual mean $E[m_0(a, L)]$ | $m(a, L)$ | $\mathbb{1}(A = a)\,/\,P(A = a \mid L)$ |
| ATE | $m(1,L) - m(0,L)$ | $A/g(L) - (1-A)/(1-g(L))$ |
| ATT numerator $E[A\{Y - m_0(0,L)\}]$ | $A\{Y - m(0,L)\}$ | $-(1-A)\,g(L)/(1-g(L))$ |
| Mean with MAR outcome $E[E(Y \mid R=1, V)]$ | $m(V)$ | $R\,/\,\pi(V)$ |
| Additive-shift MTP $E[m_0(A + \delta, L)]$ | $m(A + \delta, L)$ | $g(A - \delta \mid L)\,/\,g(A \mid L)$ |
| Subgroup mean $E[Y \mid V = v]$ | $m(v)$ | $\mathbb{1}(V = v)\,/\,P(V = v)$ |
| Sequential regime (per step $t$) | $Q_t$ evaluated under regime | product of per-step representers |

### 1.2 Why the representer de-biases

Fit $\eta_n$ with your favorite (possibly machine-learning) regression and
plug it in. Linearity gives the exact bias identity

$$
\Psi(\eta_n) - \Psi(\eta_0) = \Psi(\eta_n - \eta_0)
= E\big[\alpha_0\,(\eta_n - \eta_0)\big],
$$

so the first-order bias of the plug-in is *known up to $\alpha_0$*, and the
corrected estimator $\Psi(\eta_n) + E_n[\alpha_n\,(Y - \eta_n)]$ (the
**one-step estimator**) has only the second-order remainder

$$
E\big[(\alpha_0 - \alpha_n)(\eta_0 - \eta_n)\big],
$$

which vanishes if *either* nuisance is consistent (model double robustness)
or if the product of their convergence rates beats $n^{-1/2}$ (rate double
robustness). Concretely, the estimators in this package remain consistent if
you get the outcome regressions right but the weights wrong, or vice versa.

### 1.3 The efficient influence function

For a linear functional whose nuisance is a conditional mean
$m(X) = E[Y \mid X]$, these pieces assemble into the (uncentered) efficient
influence function

$$
\phi(O) \;=\; \underbrace{h(O;\, m)}_{\text{plug-in}} \;+\;
\underbrace{\alpha(X)\,\big(Y - m(X)\big)}_{\text{de-biasing term}},
$$

with $E[\phi] = \psi_0$. Every estimator in `{RieszCML}` is organized around
this expression, and the arguments of `RieszCurve$new()` are precisely its
components:

| Math object | `RieszCurve` argument | Example (counterfactual mean, $a=1$) |
|---|---|---|
| nuisance estimates $\eta_n$ | `nuis` | function returning `m`, `ma`, `g` |
| $\alpha(X)$ at observed data | `alpha` | `~ I(A == 1) / g` |
| $m(X)$ at observed data | `f` | `~ m` |
| $h(O; \eta)$ (under the intervention) | `h` | `~ ma` |
| uncentered EIF $\phi(O)$ | `ic_expr` | `~ h + alpha * (Y - f)` |
| $\alpha$ evaluated *under the intervention* | `alpha_star` | `~ 1 / g` |
| how TMLE updates each nuisance | `targeting_steps` | see Section 4 |

The one-step estimator is then just $n^{-1}\sum_i \phi(O_i)$
(`riesz_estimate()`), and the TMLE (`riesz_tmle()`) instead *updates the
fitted regressions* so that the plug-in itself solves
$n^{-1}\sum_i \alpha(X_i)\{Y_i - m^\star(X_i)\} = 0$, yielding a substitution
estimator that respects bounds on the outcome.

### 1.4 Nested functionals

Longitudinal regimes, CDEs, and missingness-corrected regime effects are
*sequences* of conditional expectations,

$$
\Psi(P) = E\big[h_1(\cdot\,; Q_1)\big], \qquad
Q_t = E\big[h_{t+1}(\cdot\,; Q_{t+1}) \,\big|\, \bar A_t, \bar L_t\big],
\qquad Q_T = E\big[Y \mid \bar A_T, \bar L_T\big].
$$

Each step has its own representer $\alpha_t$, and the EIF stacks them with
*cumulative products*:

$$
\phi(O) = h_1 - \psi + \sum_{t=1}^{T}
\Big(\textstyle\prod_{k \le t} \alpha_k\Big)
\big[\,h_{t+1} - Q_t\,\big].
$$

`ComposedRieszCurve` implements exactly this (Section 6), with one indexing
convention to internalize: **stages are supplied innermost-first**, so the
residual against the *innermost* regression (the one against $Y$) carries the
*full* product of representers.

## 2. Anatomy of a `RieszCurve`

We now build the counterfactual mean $E[Y^1]$ by hand, annotating every
argument. Simulated data with a known truth:

```{r}
set.seed(1)
n <- 5000

L <- rnorm(n)
A <- rbinom(n, 1, plogis(L))
Y <- rbinom(n, 1, plogis(-0.5 + 1.0 * A + 0.8 * L))

df <- data.frame(L = L, A = A, Y = Y)

# truth: E[ plogis(-0.5 + 1 + 0.8 L) ] by numerical integration
truth_mu1 <- integrate(function(l) plogis(0.5 + 0.8 * l) * dnorm(l),
                       -Inf, Inf)$value
truth_mu1
```

```{r}
rc_mu1 <- RieszCurve$new(

  ## (1) nuis: a function of `data` returning a *named list* of numeric
  ## vectors, each of length nrow(data) (scalars are recycled). This is
  ## where ALL model fitting lives. Everything the formulas below refer to
  ## must be returned here (or be a column of `data`).
  nuis = function(data) {
    m_fit <- glm(Y ~ L + A, family = binomial(), data = data)
    g_fit <- glm(A ~ L, family = binomial(), data = data)

    data1 <- data
    data1$A <- 1                      # counterfactual dataset

    list(
      m  = predict(m_fit, newdata = data,  type = "response"),  # E[Y|A,L]
      ma = predict(m_fit, newdata = data1, type = "response"),  # E[Y|1,L]
      g  = predict(g_fit, newdata = data,  type = "response")   # P(A=1|L)
    )
  },

  ## (2) alpha: the Riesz representer AT THE OBSERVED DATA, as a one-sided
  ## formula. It may reference data columns (A) and nuisance names (g).
  alpha = ~ I(A == 1) / g,

  ## (3) f: the regression at the observed data — the quantity whose
  ## residual (Y - f) the representer multiplies in the EIF.
  f = ~ m,

  ## (4) h: the plug-in transformation h(O; eta), i.e. the regression
  ## evaluated UNDER THE INTERVENTION. mean(h) is the plug-in estimator.
  h = ~ ma,

  ## (5) ic_expr: the uncentered EIF. The symbols `alpha`, `f`, `h` refer
  ## to the components above; data columns and nuisances are also visible.
  ic_expr = ~ h + alpha * (Y - f),

  ## (6) alpha_star: the representer evaluated UNDER the intervention that
  ## defines h. For alpha = 1(A=1)/g this is 1/g (set A = 1 in the
  ## indicator). Used by TMLE to update counterfactual predictions;
  ## required for composed (sequential) TMLE. Explained in Section 4.
  alpha_star = ~ 1 / g,

  ## (7) targeting_steps: how the single-curve TMLE updates each nuisance.
  ## Explained in Section 4. Ignore for one-step estimation.
  targeting_steps = list(
    m  = list(),                      # update m with the observed alpha
    ma = list(alpha_star = ~ 1 / g)   # update ma with the intervened alpha
  )
)

rc_mu1
```

Three rules about the formula interface that will save you debugging time:

1. **Formulas are evaluated in a sealed environment** containing only the
   columns of `data`, the fitted nuisance vectors, and base R functions.
   They can *not* see your global workspace, variables captured from an
   enclosing function, or non-base packages. So `~ I(A == a) / g` with `a`
   defined outside the formula **fails**; build the formula
   programmatically instead, as the catalog does:

   ```{r}
   a <- 1
   alpha_formula <- as.formula(paste0("~ I(A == ", a, ") / g"))
   alpha_formula
   ```

   Similarly, `plogis()` lives in `{stats}`, not `{base}`, so compute such
   transformations inside `nuis()` and return the result as a nuisance
   vector rather than calling them inside a formula.

2. **`h` may reference `f`**, which is convenient when the plug-in is a
   correction of the natural-value regression (e.g. `h = ~ f + delta_term`).
   `ic_expr` may reference `alpha`, `f`, and `h`.

3. **Nuisance functions must return predictions for every row.**
   `RieszCurve$fit()` rejects nuisance vectors whose length is not
   `nrow(data)` (or 1). If you fit a model on a subset — e.g. the outcome
   regression among the uncensored — fit on the subset but *predict on all
   rows*.

Fitting the curve evaluates everything and assembles the influence curve:

```{r}
rc_mu1$fit(df)

head(cbind(alpha = rc_mu1$fit_alpha, f = rc_mu1$fit_f,
           h = rc_mu1$fit_h, ic = rc_mu1$fit_ic))
mean(rc_mu1$fit_ic)   # the one-step estimate, computed by hand
```

## 3. One-step estimation: `riesz_estimate()`

`riesz_estimate()` fits the curve and returns the empirical mean of the
uncentered EIF, with a variance estimate from the sample variance of the EIF:

```{r}
fit_os <- riesz_estimate(df, rc_mu1)
fit_os
c(estimate = fit_os$estimate, truth = truth_mu1)
```

The returned `RieszFit` object carries `estimate`, `se`, `var`,
`ci_low`/`ci_high` (Wald), `n`, and `ic` (the per-observation influence
curve, which you will need for delta-method contrasts, Section 7).

The one-step estimator needs only `nuis`, `alpha`, `f`, `h`, and `ic_expr` —
`alpha_star` and `targeting_steps` are TMLE machinery. It is fast, general,
and doubly robust, but it is *not* a substitution estimator: with weak
overlap or small samples it can leave the parameter space (e.g. a negative
estimate of a risk). That is what the TMLE is for.

## 4. TMLE: `riesz_tmle()`, `targeting_steps`, and `alpha_star`

### 4.1 What the fluctuation does

TMLE turns the initial plug-in `mean(h)` into an efficient estimator by
*updating the fitted regression* rather than adding a correction term. The
update is a one-dimensional generalized regression: with the logistic
fluctuation (recommended), the outcome is rescaled to $[0, 1]$ using
`bounds`, and we fit

$$
\operatorname{logit}\, m^\star(X) \;=\; \operatorname{logit}\, m(X) \;+\;
\varepsilon\, \alpha(X),
$$

i.e. an intercept-free logistic regression of $Y$ on the **clever covariate**
$\alpha$ with the initial fit as an offset. The MLE $\hat\varepsilon$ solves
the score equation $n^{-1}\sum_i \alpha(X_i)\{Y_i - m^\star(X_i)\} = 0$ —
exactly the de-biasing term of the EIF — so the updated plug-in
$n^{-1}\sum_i h(O_i; m^\star)$ inherits the one-step estimator's asymptotics
while remaining a substitution estimator that respects `bounds`.

### 4.2 The subtlety: updating the *counterfactual* predictions

Here is the point that `targeting_steps` and `alpha_star` exist to handle,
and the single most common way to get a hand-rolled TMLE silently wrong.

The fluctuation regression is fit at the *observed* data, where the clever
covariate is $\alpha(X) = \mathbb{1}(A = 1)/g(L)$. But the final estimate is
`mean(h_star)`, which evaluates the updated regression *under the
intervention* — and under the intervention $A = 1$, the indicator is
identically 1, so the correct clever covariate for updating `ma` is
$1/g(L)$, **for every unit, including those observed with $A = 0$**. If you
instead updated `ma` with the observed-data $\alpha$, all units with
$A = 0$ would receive no update at all, and the "TMLE" would fail to be
doubly robust (its plug-in never gets repaired where repairs are needed
most). The distinction is invisible for units with $A = 1$ — which is why
this bug survives casual testing — and decisive off the intervention's
support.

`targeting_steps` declares, for each nuisance appearing in `f` or `h`, which
clever covariate updates it. Each entry is a list with at most one field:

* `list()` — update with the **observed-data** `alpha`. Correct for the
  natural-value regression `m` appearing in `f` (that is the fluctuation
  itself).
* `list(alpha_star = ~ ...)` — update with an explicit clever covariate,
  evaluated in the same (data, nuisances) environment as the other
  formulas. **This is the recommended, fully general mechanism.** For
  indicator representers it is `~ 1/g`; for missingness weights `R/pi` it
  is `~ 1/pi`; for a density-ratio (shift) representer it is the ratio
  re-evaluated at the shifted exposure (Section 5.3).
* `list(set = list(A = 1))` — re-evaluate the `alpha` formula on data with
  the listed columns overwritten. Convenient when `alpha` is written
  directly in terms of data columns (e.g. `~ I(A == 1)/g`), but *invalid*
  when `alpha` is built purely from precomputed nuisance vectors (a density
  ratio does not change when you overwrite a column it never reads). Prefer
  `alpha_star`.

All updates share the single fluctuation coefficient $\hat\varepsilon$; they
differ only in the covariate that multiplies it.

The curve-level `alpha_star` argument plays the analogous role for each
*stage* of a composed TMLE (Section 6.3), where per-stage `targeting_steps`
are not consulted.

### 4.3 Running the TMLE

```{r}
fit_tmle <- riesz_tmle(
  data = df,
  rc = rc_mu1,
  fluctuation_type = "logistic",  # "logistic" (recommended) or "identity"
  bounds = c(0, 1),               # known range of Y; required for logistic
  outcome_col = "Y"               # the variable the fluctuation regresses
)

fit_tmle
c(one_step = fit_os$estimate, tmle = fit_tmle$estimate, truth = truth_mu1)
```

Arguments worth knowing:

* `fluctuation_type = "logistic"` performs the update on the
  logit scale after mapping the outcome into $[0,1]$ via `bounds`; the
  updated predictions are guaranteed to respect `bounds`. Use it for
  binary outcomes with `bounds = c(0, 1)` and for continuous outcomes with
  `bounds = range(Y)` (possibly padded). `"identity"` fits a linear
  fluctuation and needs no bounds, but offers no range protection.
* `outcome_col` names the fluctuation target. For a single curve it should
  be the variable appearing in the residual of your `ic_expr`.
* `fluctuation_weights` passes observation weights into the fluctuation
  regression (e.g. inverse-probability-of-censoring weighted TMLEs).
* `score_tol`: after a successful update, the mean of the *targeted*
  influence curve must equal the estimate up to numerical error;
  `riesz_tmle()` warns when it does not. Treat that warning as "my
  `targeting_steps`/`alpha_star` are wrong", not as noise.

The returned fit contains both `ic` (the untargeted EIF used for the
variance and confidence intervals) and `ic_star` (the targeted influence
curve used for the self-check), along with `eps`, `f_star`, `h_star`, and
the fluctuation model(s).

```{r}
# Diagnostic: the targeted score equation is solved
c(mean_ic_star = mean(fit_tmle$ic_star), estimate = fit_tmle$estimate)

# The updated predictions respect the bounds
range(fit_tmle$h_star)
```

### 4.4 Why bother? A stress test

With poor initial regressions and near-positivity problems, the one-step
estimator can escape $[0, 1]$; the bounded TMLE cannot, and both remain
consistent as long as the weights are consistent (double robustness):

```{r}
set.seed(6)
n2 <- 150
L2 <- rnorm(n2)
A2 <- rbinom(n2, 1, plogis(-2.5 + 4.5 * L2))        # strong confounding
Y2 <- rbinom(n2, 1, plogis(-1 + 2 * L2 + 1.5 * A2))
df2 <- data.frame(L = L2, A = A2, Y = Y2)
truth2 <- integrate(function(l) plogis(0.5 + 2 * l) * dnorm(l),
                    -Inf, Inf)$value

rc_bad <- riesz_curve_catalog$cfmean_a(
  a = 1,
  nuis = function(data) {
    m_fit <- lm(Y ~ L + A, data = data)   # deliberately crude linear fit
    g_fit <- glm(A ~ L, family = binomial(), data = data)
    d1 <- data; d1$A <- 1
    list(
      m  = predict(m_fit, newdata = data),
      ma = predict(m_fit, newdata = d1),
      g  = predict(g_fit, newdata = data, type = "response")
    )
  }
)

c(one_step = riesz_estimate(df2, rc_bad)$estimate,   # a "probability"...
  tmle = riesz_tmle(df2, rc_bad, fluctuation_type = "logistic",
                    bounds = c(0, 1), outcome_col = "Y")$estimate,
  truth = truth2)
```

A handful of huge weights are enough to drag the one-step estimator far
outside $[0, 1]$; the TMLE processes the same information through a bounded
update of the regression and cannot leave the parameter space.

Here we used the catalog constructor `riesz_curve_catalog$cfmean_a()`,
which builds exactly the curve of Section 2 — including its
`targeting_steps` — from a `nuis` function alone. 

## 5. A recipe for estimands not in the catalog

To implement a new estimand, walk through six questions:

1. **Write the parameter as $\Psi(\eta) = E[h(O; \eta)]$.** What is the
   nuisance $\eta$ (usually a regression)? What is the linear map $h$?
2. **Find the representer by change of measure.** Write the parameter as
   $\int \eta \, dP^\star$ and read off $\alpha = dP^\star/dP$: indicators
   over propensities for static regimes, $R/\pi$ for missingness, density
   ratios for shifts, and products of these for nested/coarsened problems.
3. **Write `nuis()`.** Fit every model once; return the fitted values *at
   the observed data* and *under the intervention*, plus every weight
   component your formulas need.
4. **Fill in `alpha`, `f`, `h`, `ic_expr`.** For conditional-mean nuisances,
   `ic_expr = ~ h + alpha * (Y - f)` (with `Y` replaced by whatever the
   regression's outcome is).
5. **For TMLE, derive `alpha_star`** — the representer with the intervention
   substituted into it — and declare `targeting_steps` for every nuisance in
   `f` and `h`.
6. **Validate.** Check `mean(fit_ic)` against a simulation truth; check
   double robustness by deliberately misspecifying one nuisance at a time;
   check that `mean(ic_star)` matches the TMLE estimate (the package warns
   otherwise).

The remainder of the vignette applies this recipe to progressively harder
estimands: the ATT (a *ratio* of linear functionals), a missingness-corrected
mean (where `alpha_star` earns its keep), an MTP (a density-ratio
representer), and nested functionals (longitudinal regimes and the CDE).

### 5.1 ATT: a smooth transformation of linear functionals

The ATT
$\psi = E[Y^1 - Y^0 \mid A = 1] = E[A\{Y - m_0(0, L)\}] \,/\, P(A = 1)$
is not itself linear — it is a **ratio** of two linear functionals. The
package's pattern for such parameters: fit each linear component as its own
curve, then combine with the delta method (Section 7). The numerator
functional $\theta = E[A\{Y - m_0(0,L)\}]$ has $h(O) = A\{Y - m(0,L)\}$ and
representer $\alpha(O) = -(1 - A)\, g(L)/\{1 - g(L)\}$; the denominator
$P(A=1)$ is the trivial functional with $\phi = A$. `riesz_att()` wires the
two together:

```{r}
set.seed(3)
n3 <- 5000
L3 <- rnorm(n3)
A3 <- rbinom(n3, 1, plogis(L3))
Y3 <- 2 + L3 + 1.5 * A3 + rnorm(n3, sd = 0.25)   # constant effect: ATT = 1.5
df3 <- data.frame(L = L3, A = A3, Y = Y3)

fit_att <- riesz_att(df3, nuis = function(data) {
  g_fit  <- glm(A ~ L, family = binomial(), data = data)
  m0_fit <- lm(Y ~ L, data = subset(data, A == 0))  # fit on controls...
  list(
    m0 = predict(m0_fit, newdata = data),           # ...predict on everyone
    g  = predict(g_fit, newdata = data, type = "response")
  )
})
fit_att
```

First, any parameter $b(\psi_1, \ldots, \psi_k)$ with
smooth $b$ and linear components is in reach: estimate each component, then
`riesz_delta()`. Second, note that `riesz_att()` is one-step only: because
$h$ for the numerator involves $Y$ itself rather than a counterfactual
regression evaluation, there is no `alpha_star`-style plug-in update, and
`riesz_curve_catalog$att_theta` accordingly declares no `targeting_steps`.

### 5.2 Missing outcomes: where `alpha_star` earns its keep

Under outcome missingness at random given $V$, the mean
$E[Y] = E[\,m_0(V)\,]$ with $m_0(V) = E[Y \mid R = 1, V]$ has representer
$\alpha = R/\pi(V)$, $\pi(V) = P(R = 1 \mid V)$. Here `h = ~ m` equals
`f = ~ m` — there is no counterfactual dataset — but the *update* of the
plug-in must reach the units with $R = 0$: the clever covariate for `m` in
`targeting_steps` is $1/\pi(V)$ (all units), while the fluctuation score
uses $R/\pi(V)$ (only observed residuals are informative). The catalog entry
encodes this:

```{r}
set.seed(4)
n4 <- 20000
L4 <- rnorm(n4)
A4 <- rbinom(n4, 1, plogis(L4))
Yf <- 2 + L4 + 1.5 * A4 + rnorm(n4, sd = 0.25)
R4 <- rbinom(n4, 1, plogis(-0.5 + 0.5 * L4 + 0.3 * A4))
df4 <- data.frame(L = L4, A = A4, R = R4, Y = ifelse(R4 == 1, Yf, 0))
truth_mean <- 2 + 1.5 * 0.5   # E[L]=0, E[A]=1/2

rc_miss <- riesz_curve_catalog$missing_mean(nuis = function(data) {
  obs <- data[data$R == 1, , drop = FALSE]
  pi_fit <- glm(R ~ L + A, family = binomial(), data = data)
  list(
    m  = rep(mean(obs$Y), nrow(data)),   # deliberately misspecified!
    pi = predict(pi_fit, newdata = data, type = "response")
  )
})

# the naive plug-in is biased ...
rc_miss$fit(df4); c(plugin = mean(rc_miss$fit_h), truth = truth_mean)

# ... the TMLE repairs it, because 1/pi updates every unit
fit_miss <- riesz_tmle(df4, rc_miss, fluctuation_type = "logistic",
                       bounds = range(df4$Y) + c(-1, 1), outcome_col = "Y")
c(tmle = fit_miss$estimate, truth = truth_mean)
```

Had `missing_mean` updated `m` with the observed covariate `R/pi`, the
intercept-only regression would be repaired only among the observed, and
this display would show a biased TMLE. When you write your own censoring or
sampling corrections, this is the detail to check first.

### 5.3 Modified treatment policies: density-ratio representers

For a continuous exposure and the additive shift $d(A, L) = A + \delta$, the
target $E[\,m_0(A + \delta, L)\,]$ has, by change of measure of the exposure
density,

$$
\alpha(A, L) = \frac{g(A - \delta \mid L)}{g(A \mid L)},
\qquad
\alpha^\star(A, L) = \alpha(A + \delta, L)
                   = \frac{g(A \mid L)}{g(A + \delta \mid L)},
$$

where $\alpha^\star$ — the representer *evaluated at the shifted exposure* —
is the clever covariate for updating the shifted predictions `m_shift`. Note
that `targeting_steps = list(m_shift = list(set = ...))` would be **wrong**
here: `alpha` is computed from precomputed density columns, so overwriting
`A` in the data changes nothing. This is exactly why the general
`alpha_star` mechanism exists, and why `riesz_curve_catalog$mtp()` requires
you to supply it for TMLE:

```{r}
set.seed(5)
n5 <- 5000
delta <- 0.5
L5 <- rnorm(n5)
A5 <- rnorm(n5, mean = 0.5 * L5, sd = 1)          # A | L ~ N(0.5 L, 1)
Y5 <- 1 + 1.2 * A5 + 0.8 * L5 + rnorm(n5, sd = 0.5)
df5 <- data.frame(L = L5, A = A5, Y = Y5)
truth_mtp <- 1 + 1.2 * delta                      # E[Y] shifts by beta_A * delta

rc_mtp <- riesz_curve_catalog$mtp(
  nuis = function(data) {
    m_fit <- lm(Y ~ A + L, data = data)
    d_shift <- data; d_shift$A <- d_shift$A + delta

    # parametric conditional density of A | L (Gaussian working model)
    a_fit <- lm(A ~ L, data = data)
    mu <- predict(a_fit, newdata = data)
    s  <- summary(a_fit)$sigma

    list(
      m       = predict(m_fit, newdata = data),
      m_shift = predict(m_fit, newdata = d_shift),
      g       = dnorm(data$A,         mean = mu, sd = s),  # g(A | L)
      g_shift = dnorm(data$A - delta, mean = mu, sd = s),  # g(A - d | L)
      g_up    = dnorm(data$A + delta, mean = mu, sd = s)   # g(A + d | L)
    )
  },
  alpha_star = ~ g / g_up    # representer at the shifted exposure
)

fit_mtp_os <- riesz_estimate(df5, rc_mtp)
fit_mtp <- riesz_tmle(df5, rc_mtp, fluctuation_type = "logistic",
                      bounds = range(df5$Y) + c(-1, 1), outcome_col = "Y")
c(one_step = fit_mtp_os$estimate, tmle = fit_mtp$estimate, truth = truth_mtp)
```

With a flexible density-ratio estimator, replace the Gaussian working model
inside `nuis()` — e.g. the probabilistic-classifier trick of Díaz et al.
(2021), in which a classifier is trained to distinguish observed from
shifted exposures and the ratio is recovered as ` r / (1 - r)`; the package
never sees how the ratio was produced, only the resulting vectors.

## 6. Nested functionals: `ComposedRieszCurve`

### 6.1 Conventions

A `ComposedRieszCurve` chains single-stage curves to implement the
sequential-regression EIF of Section 1.4. The conventions:

* **Stages are supplied innermost-first.** `rc_list[[1]]` is the regression
  fit directly against the outcome `Y` — the *last* time point. Each later
  stage regresses the previous stage's `h`, which it receives as a data
  column literally named `` `h[j=1]` ``, `` `h[j=2]` ``, and so on (note the
  backticks in formulas).
* **Residual weights are suffix products.** Stage $j$'s residual is weighted
  by $\omega_j = \prod_{k \ge j} \alpha_k$: the products accumulate from the
  *outermost* stage (first time point) inward, so the residual against the
  innermost regression, $Y - f_1$, carries the full product of all
  representers, while the outermost residual carries that stage's
  representer alone. The class computes this for you (`fit_omega`), but you
  should know it when checking an EIF by hand — accumulating the products
  the other way preserves consistency under correct specification while
  silently destroying double robustness.
* **Each stage intervenes only on its own node.** Stage $j$'s `h` evaluates
  its regression with only its own treatment set; earlier time points stay
  at their natural values (they are handled by the later stages). This is
  the standard sequential g-formula recursion.

### 6.2 Longitudinal regime mean $E[Y^{1,1}]$

Two time points with treatment–confounder feedback; the truth is
$E[Y^{1,1}] = 7.75$ in this design:

```{r}
set.seed(6)
sim_long <- function(n) {
  L1 <- rnorm(n)
  A1 <- rbinom(n, 1, plogis(-0.2 + 1.0 * L1))
  L2 <- 0.5 + 0.7 * L1 + 2.0 * A1 + rnorm(n)
  A2 <- rbinom(n, 1, plogis(-0.3 + 0.6 * A1 + 1.2 * L2))
  Y  <- 1.0 + 1.5 * L2 + 3.0 * A2 + rnorm(n)
  data.frame(L1 = L1, A1 = A1, L2 = L2, A2 = A2, Y = Y)
}
df6 <- sim_long(10000)
truth_mu11 <- 1.0 + 1.5 * (0.5 + 2.0) + 3.0

## innermost stage (time 2): regress Y, intervene on A2
rc_t2 <- RieszCurve$new(
  nuis = function(data) {
    m_fit <- lm(Y ~ L1 + A1 + L2 + A2, data = data)
    g_fit <- glm(A2 ~ L1 + A1 + L2, family = binomial(), data = data)
    d1 <- data; d1$A2 <- 1
    list(m  = predict(m_fit, newdata = data),
         m1 = predict(m_fit, newdata = d1),
         g  = predict(g_fit, newdata = data, type = "response"))
  },
  f = ~ m, h = ~ m1,
  alpha = ~ A2 / g,
  alpha_star = ~ 1 / g,                    # required for composed TMLE
  ic_expr = ~ h + alpha * (Y - f)
)

## outermost stage (time 1): regress the previous h, intervene on A1
rc_t1 <- RieszCurve$new(
  nuis = function(data) {
    m_fit <- lm(`h[j=1]` ~ L1 + A1, data = data)
    g_fit <- glm(A1 ~ L1, family = binomial(), data = data)
    d1 <- data; d1$A1 <- 1
    list(m  = predict(m_fit, newdata = data),
         m1 = predict(m_fit, newdata = d1),
         g  = predict(g_fit, newdata = data, type = "response"))
  },
  f = ~ m, h = ~ m1,
  alpha = ~ A1 / g,
  alpha_star = ~ 1 / g,
  ic_expr = ~ h + alpha * (`h[j=1]` - f)
)

rc_11 <- ComposedRieszCurve$new(rc_list = list(rc_t2, rc_t1))

fit_11_os <- riesz_estimate(df6, rc_11)
fit_11_tmle <- riesz_tmle(df6, rc_11, fluctuation_type = "logistic",
                          bounds = range(df6$Y) + c(-1, 1), outcome_col = "Y")

c(one_step = fit_11_os$estimate,
  tmle = fit_11_tmle$estimate,
  truth = truth_mu11)
```

For an **LMTP** — a longitudinal modified treatment policy — the recipe is
identical, except each stage uses a density-ratio `alpha` (and matching
shifted-exposure `alpha_star`) exactly as in Section 5.3, in place of the
indicator-over-propensity representer.

### 6.3 How the composed TMLE targets

`riesz_tmle()` on a composed curve runs the sequential-regression TMLE:
proceeding innermost-first, it fluctuates stage $j$'s `f` against the current
target (first `Y`, thereafter the previous stage's updated `h`) with clever
covariate $\omega_j = \prod_{k \ge j} \alpha_k$, then updates stage $j$'s
counterfactual predictions with the *intervened* covariate
$\omega_j^\star = \alpha_j^\star \prod_{k > j} \alpha_k$ — the same
observed-versus-intervened distinction as in Section 4.2, applied stagewise.
Consequently **every stage of a composed TMLE must supply `alpha_star`**;
the per-stage `targeting_steps` are not consulted in the composed path. A
stage without `alpha_star` triggers a loud warning and an update that is not
a valid TMLE in general.

The estimator is doubly robust in the classical sequential sense:
consistency follows if all the sequential regressions or all the
representers are consistently estimated.

### 6.4 Controlled direct effects: same skeleton, different labels

The CDE mean $E[Y^{a, m}]$ — outcome under a *joint* static intervention on
treatment $A$ and mediator $M$ — is, via the sequential g-formula,

$$
\psi_{a,m} = E\Big[\, E\big[\, E[Y \mid L, A, M]\big|_{M=m} \,\Big|\, L, A\big]\Big|_{A=a}\, \Big],
$$

with representers $\alpha_2 = \mathbb{1}(M = m)/P(M = m \mid A, L)$
(innermost) and $\alpha_1 = \mathbb{1}(A = a)/P(A = a \mid L)$. Structurally
this is the two-time-point regime with $(A_1, A_2)$ relabeled $(A, M)$:

```{r}
set.seed(7)
n7 <- 10000
L7 <- rnorm(n7)
A7 <- rbinom(n7, 1, plogis(L7))
M7 <- rbinom(n7, 1, plogis(-0.5 + 0.5 * L7 + 1.0 * A7))
Y7 <- 1 + L7 + 2 * A7 + 1.5 * M7 + 1.0 * A7 * M7 + rnorm(n7)
df7 <- data.frame(L = L7, A = A7, M = M7, Y = Y7)
truth_cde11 <- 1 + 2 + 1.5 + 1.0     # E[Y^{a=1,m=1}] = 5.5

rc_M <- RieszCurve$new(              # innermost: intervene on M
  nuis = function(data) {
    m_fit <- lm(Y ~ L + A * M, data = data)
    g_fit <- glm(M ~ L + A, family = binomial(), data = data)
    d1 <- data; d1$M <- 1
    list(m  = predict(m_fit, newdata = data),
         m1 = predict(m_fit, newdata = d1),
         g  = predict(g_fit, newdata = data, type = "response"))
  },
  f = ~ m, h = ~ m1,
  alpha = ~ I(M == 1) / g, alpha_star = ~ 1 / g,
  ic_expr = ~ h + alpha * (Y - f)
)

rc_A <- RieszCurve$new(              # outermost: intervene on A
  nuis = function(data) {
    m_fit <- lm(`h[j=1]` ~ L + A, data = data)
    g_fit <- glm(A ~ L, family = binomial(), data = data)
    d1 <- data; d1$A <- 1
    list(m  = predict(m_fit, newdata = data),
         m1 = predict(m_fit, newdata = d1),
         g  = predict(g_fit, newdata = data, type = "response"))
  },
  f = ~ m, h = ~ m1,
  alpha = ~ I(A == 1) / g, alpha_star = ~ 1 / g,
  ic_expr = ~ h + alpha * (`h[j=1]` - f)
)

rc_cde11 <- ComposedRieszCurve$new(rc_list = list(rc_M, rc_A))

fit_cde11 <- riesz_tmle(df7, rc_cde11, fluctuation_type = "logistic",
                        bounds = range(df7$Y) + c(-1, 1), outcome_col = "Y")
c(estimate = fit_cde11$estimate, truth = truth_cde11)
```

The CDE contrast $E[Y^{1,m}] - E[Y^{0,m}]$ is then a delta-method difference
of two such fits (next section). Natural direct effects use the same nested
skeleton with a mediator-density-*ratio* representer,
$P(M \mid A = 0, L)/P(M \mid A, L)$, in place of the indicator — i.e. the
mediation analogue of an MTP stage.

Composition also mixes problem types freely: prepend a `missing_mean`-style
stage to correct any of the above for a MAR outcome, giving, e.g., a
longitudinal regime effect with a censored final outcome as a three-stage
composed curve.

## 7. Contrasts and other smooth transformations: `riesz_delta_*()`

Any smooth function $b(\psi_1, \ldots, \psi_k)$ of fitted linear functionals
is estimated by the delta method from the components' influence curves. Fit
each component (all fits must be on the *same data*, so the influence curves
align row-wise), then combine:

```{r}
rc_mu0 <- riesz_curve_catalog$cfmean_a(
  a = 0,
  nuis = function(data) {
    m_fit <- glm(Y ~ L + A, family = binomial(), data = data)
    g_fit <- glm(I(A == 0) ~ L, family = binomial(), data = data)
    d0 <- data; d0$A <- 0
    list(m  = predict(m_fit, newdata = data, type = "response"),
         ma = predict(m_fit, newdata = d0, type = "response"),
         g  = predict(g_fit, newdata = data, type = "response"))
  }
)

fit_mu0 <- riesz_tmle(df, rc_mu0, fluctuation_type = "logistic",
                      bounds = c(0, 1), outcome_col = "Y")

riesz_delta_difference(fit_tmle, fit_mu0, parameter = "ATE = E[Y^1] - E[Y^0]")
riesz_delta_ratio(fit_tmle, fit_mu0, parameter = "Risk ratio")
```

Also available: `riesz_delta_log_ratio()`, `riesz_delta_odds_ratio()`,
`riesz_delta_log_odds_ratio()`, and the general `riesz_delta(fits =, g =,
grad =)` for arbitrary smooth `g` (the gradient is meant to be supplied by the 
user as a function of the parameter). This is how one should approach ATEs of composed curves,
regime contrasts, CDE differences, ratios of MTP means, the ATT, and many other parameters.

## 8. Diagnostics, pitfalls, and good habits

**Check the targeted score.** After `riesz_tmle()`, `mean(fit$ic_star)`
should equal `fit$estimate` to numerical precision; the function warns via
`score_tol` when it does not. The usual culprits are a missing or wrong
`alpha_star`, `set =` used with a nuisance-built `alpha`, or `bounds` that
clip the data.

**Stress-test double robustness in simulation.** Before trusting a new
curve, deliberately break one nuisance at a time (e.g. replace a regression
with its intercept-only version) and confirm the estimator still recovers a
known truth. This is the test that catches the two classic sequential-TMLE
bugs — reversed cumulative products and observed-covariate counterfactual
updates — both of which are invisible when everything is correctly
specified.

**Formulas are sealed.** No globals, no closure variables, no non-base
functions. Build formulas with `paste0()`/`as.formula()` when they depend on
parameters like the treatment level or shift size, and do all nontrivial
computation inside `nuis()`.

**Nuisances must cover every row.** Fit on subsets when the estimand demands
it (controls only, uncensored only), but always `predict(..., newdata =
data)` on the full data; `fit()` errors on length mismatches and `NA`s
rather than allowing silent misalignment.

**Positivity violations can lead to unstable estimators.** Representers are
weights; inspect `rc$fit_alpha` (and `fit_omega` for composed curves) for
extreme values. For long time series, the manuscript notes that the
cumulative product $\omega_t = \prod_{k \le t} \alpha_k$ is itself a Riesz
representer and may be estimated *directly* (e.g. by Riesz regression) for
stability — the composed interface accepts whatever vectors `nuis()`
returns, so nothing stops you from supplying $\omega_t$ estimates in a
custom `ic_expr`.

**Cross-fit adaptive learners.** The asymptotics require either Donsker
conditions or sample splitting. `nuis()` is agnostic about how predictions
are produced, so out-of-fold predictions from a cross-fitted Super Learner
drop straight in; see the companion vignette *Cross-fitted Super Learning
with nadir*.

**Inference notes.** Confidence intervals are Wald intervals based on the
sample variance of the estimated (untargeted) EIF; this is the standard
first-order approximation and can be optimistic when nuisances are
inconsistently estimated or overlap is weak.

## 9. Argument reference card

| Argument | Used by | Meaning |
|---|---|---|
| `nuis` | everything | function of `data` returning a named list of prediction vectors (length `nrow(data)` or 1); or the precomputed list itself |
| `alpha` | EIF, one-step, TMLE fluctuation | Riesz representer at the observed data (formula over data columns + nuisances, or numeric) |
| `f` | EIF, TMLE offset | regression at the observed data; the residual target |
| `h` | plug-in, EIF | $h(O;\eta)$: regression evaluated under the intervention; may reference `f` |
| `ic_expr` | one-step, diagnostics | uncentered EIF; sees `alpha`, `f`, `h`, data, nuisances |
| `alpha_star` | composed TMLE (per stage); `mtp` catalog | representer evaluated under the intervention defining `h` |
| `targeting_steps` | single-curve TMLE | per-nuisance update spec: `list()` = observed `alpha`; `list(alpha_star = ~ ...)` = explicit clever covariate (recommended); `list(set = list(col = val))` = re-evaluate `alpha` on intervened data (indicator representers only) |
| `fluctuation_type`, `bounds` | `riesz_tmle()` | `"logistic"` + outcome bounds (recommended; range-respecting) or `"identity"` |
| `outcome_col` | `riesz_tmle()` | the fluctuation target; the variable in your innermost residual |
| `fluctuation_weights`, `clip`, `score_tol` | `riesz_tmle()` | fluctuation observation weights; logit-scale clipping; targeted-score warning threshold |

## References

* Balkus, S. V., Testa, C., and Hejazi, N. S. (2026). A Riesz representer
  perspective on targeted learning. (The accompanying manuscript; Theorem 1
  gives the single-step Riesz EIF, Theorem 2 the sequential EIF implemented
  by `ComposedRieszCurve`, and Algorithms 1 and 3 the TMLEs implemented by
  `riesz_tmle()`.)
* Chernozhukov, V., Newey, W. K., and Singh, R. (2022). Automatic debiased
  machine learning of causal and structural effects. *Econometrica*.
* Díaz, I., Williams, N., Hoffman, K. L., and Schenck, E. J. (2021).
  Nonparametric causal effects based on longitudinal modified treatment
  policies. *JASA*.
* Hines, O., Dukes, O., Diaz-Ordaz, K., and Vansteelandt, S. (2022).
  Demystifying statistical learning based on efficient influence functions.
  *The American Statistician*.
* van der Laan, M. J. and Gruber, S. (2012). Targeted minimum loss based
  estimation of causal effects of multiple time point interventions.
  *The International Journal of Biostatistics*.
* van der Laan, M. J. and Rose, S. (2011). *Targeted Learning*. Springer.
* Williams, N. T., Hines, O. J., and Rudolph, K. E. (2025). Riesz
  representers for the rest of us.
