---
title: "Cross-fitted Super Learning with nadir"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-fitted Super Learning with nadir}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
# the version of nadir with crossfit_super_learner() is not (yet) on CRAN
library(nadir)


knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`{RieszCML}` is agnostic about how nuisance functions are estimated: the
`nuis` function of a `RieszCurve` just has to return predictions for every
row of `data`. This vignette shows how to plug in `{nadir}`'s Super Learner,
first without and then with cross-fitting, using
`nadir::crossfit_super_learner()`.

Why cross-fit? The asymptotic guarantees of one-step estimators and TMLEs
require the empirical process term to vanish, which holds either under
Donsker (complexity) conditions on the nuisance estimators or, without any
such conditions, when each observation's nuisance predictions come from
models fit without that observation. Adaptive learners like random forests
and gradient boosting can violate Donsker conditions, so cross-fitting is the
safe default when they are in the library. `crossfit_super_learner()`
implements this properly: on each outer training split it fits a *full*
Super Learner — so the ensemble weights, not just the base learners, are
estimated without the held-out fold — and returns out-of-fold predictions in
the original row order, which is exactly what a `nuis` function must supply.

```{r}
library(RieszCML)
library(nadir)

simulate_data <- function(n, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)

  L1 <- rnorm(n)
  L2 <- rnorm(n)
  L3 <- rnorm(n)
  L4 <- rnorm(n)

  eta_A <- -0.4 +
    1.2 * sin(L1) -
    0.8 * L2^2 +
    0.7 * L3 * L4 +
    0.5 * (L1 > 0) -
    0.5 * (L2 > 1)

  g <- expit(eta_A)
  A <- rbinom(n, size = 1, prob = g)

  eta_Y <- -0.8 +
    0.8 * A +
    0.6 * A * L1 -
    0.5 * A * (L2 > 0) +
    0.7 * sin(L1 * L3) +
    0.4 * L2^2 -
    0.3 * abs(L4)

  Q <- expit(eta_Y)
  Y <- rbinom(n, size = 1, prob = Q)

  data.frame(L1 = L1, L2 = L2, L3 = L3, L4 = L4, A = A, Y = Y)
}

truth_ate_mc <- function(n_mc = 2e6, seed = 123) {
  df_mc <- simulate_data(n_mc, seed = seed)

  eta_Y1 <- with(df_mc,
    -0.8 + 0.8 + 0.6 * L1 - 0.5 * (L2 > 0) +
      0.7 * sin(L1 * L3) + 0.4 * L2^2 - 0.3 * abs(L4)
  )
  eta_Y0 <- with(df_mc,
    -0.8 + 0.7 * sin(L1 * L3) + 0.4 * L2^2 - 0.3 * abs(L4)
  )

  mean(plogis(eta_Y1) - plogis(eta_Y0))
}

truth_ate <- truth_ate_mc()
truth_ate
```

Both analyses share the same working formulas and learner library.

```{r}
outcome_formula <- Y ~ A + L1 + L2 + L3 + L4 +
  I(L1 * L3) + I(L2^2) + I(abs(L4)) + A:L1 + A:I(L2 > 0)

prop_formula <- A ~ L1 + L2 + L3 + L4 +
  I(sin(L1)) + I(L2^2) + I(L3 * L4) + I(L1 > 0) + I(L2 > 1)

learner_library <- list(
  logistic = nadir::lnr_logistic,
  rf       = nadir::lnr_rf_binary,
  xgboost  = nadir::lnr_xgboost,
  glmnet   = nadir::lnr_glmnet
)

set.seed(1)
df <- simulate_data(4000, seed = 1)
```

## Super Learner nuisances without cross-fitting

The simplest approach fits one Super Learner to all of the data and predicts
in-sample. This is fine when the library contains only well-behaved
(Donsker) learners, and serves as our baseline.

```{r}
make_nuis_superlearner_nocf <- function() {
  function(data) {
    m_fit <- nadir::super_learner(
      data = data,
      formula = outcome_formula,
      learners = learner_library,
      outcome_type = "binary"
    )

    g_fit <- nadir::super_learner(
      data = data,
      formula = prop_formula,
      learners = learner_library,
      outcome_type = "binary"
    )

    data1 <- data
    data0 <- data
    data1$A <- 1
    data0$A <- 0

    list(
      m  = predict(m_fit, data),
      m1 = predict(m_fit, data1),
      m0 = predict(m_fit, data0),
      g  = predict(g_fit, data)
    )
  }
}

fit_ate_nocf <- riesz_tmle(
  data = df,
  rc = riesz_curve_catalog$ate(nuis = make_nuis_superlearner_nocf()),
  fluctuation_type = "logistic",
  bounds = c(0, 1),
  outcome_col = "Y"
)

fit_ate_nocf
c(estimate = fit_ate_nocf$estimate, truth = truth_ate)
```

## Cross-fitted nuisances with `crossfit_super_learner()`

`nadir::crossfit_super_learner()` handles the entire cross-fitting protocol
that one would otherwise write by hand: it splits the data into `n_folds`
outer folds, fits a full Super Learner (with its own `inner_n_folds`-fold
internal CV for the ensemble weights) on each outer training split, and
exposes the results through three accessors that map directly onto the
nuisances a `RieszCurve` needs:

* `$oof_predictions()` — out-of-fold predictions at the observed data, in
  original row order: this is `m` (or `g`);
* `$predict_modified(modify)` — out-of-fold predictions after applying an
  intervention to each held-out fold, e.g. setting `A = 1`: this is `m1`;
* `$sl_fits`, `$cv_loss`, `$fold_assignments` — diagnostics, used below.

```{r}
make_nuis_superlearner_cf <- function(n_folds = 5, inner_n_folds = 5) {
  function(data) {
    cf_m <- nadir::crossfit_super_learner(
      data = data,
      formulas = outcome_formula,
      learners = learner_library,
      n_folds = n_folds,
      inner_n_folds = inner_n_folds,
      outcome_type = "binary"
    )

    cf_g <- nadir::crossfit_super_learner(
      data = data,
      formulas = prop_formula,
      learners = learner_library,
      n_folds = n_folds,
      inner_n_folds = inner_n_folds,
      outcome_type = "binary"
    )

    list(
      m  = cf_m$oof_predictions(),
      m1 = cf_m$predict_modified(function(d) { d$A <- 1; d }),
      m0 = cf_m$predict_modified(function(d) { d$A <- 0; d }),
      g  = cf_g$oof_predictions()
    )
  }
}

set.seed(1)
fit_ate_cf <- riesz_tmle(
  data = df,
  rc = riesz_curve_catalog$ate(nuis = make_nuis_superlearner_cf()),
  fluctuation_type = "logistic",
  bounds = c(0, 1),
  outcome_col = "Y"
)

fit_ate_cf
c(estimate = fit_ate_cf$estimate, truth = truth_ate)
```

Compared to a hand-rolled fold loop, note what came for free here: the
ensemble weights of each fold's Super Learner were estimated without the
held-out observations (a hand-rolled loop that calls `super_learner()` per
fold gets this too, but only if written carefully); the validation folds are
checked to be disjoint; predictions are reconstructed in original row order;
and if any row were never held out, its prediction would be `NA` — which
`RieszCurve$fit()` rejects loudly rather than letting a silently misaligned
nuisance vector through.

## Inspecting the cross-fit

Because the fitted objects are retained, the cross-fit is easy to audit. It
can be worth doing this *outside* the `nuis` function during exploratory
work, then wrapping the final recipe as above:

```{r}
set.seed(1)
cf_m <- nadir::crossfit_super_learner(
  data = df,
  formulas = outcome_formula,
  learners = learner_library,
  n_folds = 5,
  outcome_type = "binary"
)

cf_m

# honest empirical loss of the cross-fitted outcome predictions
cf_m$cv_loss

# ensemble weights estimated within each outer training split
lapply(cf_m$sl_fits, `[[`, "learner_weights")

# which outer fold held out each observation
table(cf_m$fold_assignments)
```

Stable learner weights across folds suggest the ensemble is not being driven
by a few influential observations; wildly varying weights are a hint to
enlarge the library or the sample.

## Practical notes

**Sharing folds across nuisances.** As written, `m` and `g` are cross-fitted
on independently drawn fold splits, which is standard and asymptotically
valid. If you prefer that both nuisances share the same outer folds (e.g.
for exact reproducibility of fold-level diagnostics), pass a common
`cv_schema` function to both `crossfit_super_learner()` calls.

**Parallelism.** Fitting is parallelized over outer folds via the `{future}`
framework: load `{future}` and declare, e.g., `plan(multisession)` before
calling `crossfit_super_learner()`; use
`options(future.globals.maxSize = Inf)` if large learner objects need to be
exported to workers. A nested topology (see `?future::plan`) additionally
parallelizes the inner Super Learners. This vignette runs sequentially so it
builds anywhere.

**Cross-fitting the fluctuation?** `riesz_tmle()` fits its one-dimensional
fluctuation on the full sample. This is the standard CV-TMLE-adjacent
construction: with cross-fitted initial nuisances, the parametric
fluctuation step does not reintroduce Donsker-type problems.
