---
title: "Average derivative effects in subgroups"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Average derivative effects in subgroups}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview 

This vignette illustrates how the RieszCML framework can represent an average derivative effect (ADE) within a subgroup.

The target parameter is
$$\psi = \mathbb{E}\left[
\frac{\partial}{\partial a} \mathbb{E}[Y \mid A = a, L]
\;\middle|\;
V = 1
\right],
$$
where $V$ defines a subgroup of interest. 
Equivalently, 
$$\psi
=
\frac{1}{P(V=1)}
\mathbb{E}\left[
I(V=1)
\frac{\partial}{\partial a} m(A,L)
\right],$$
$$m(a,L) = \mathbb{E}[Y \mid A=a, L].$$
The Riesz representer is 
$$\alpha(O)
=
\frac{I(V=1)}{P(V=1)}
\frac{\partial}{\partial a}\log p(A \mid L),$$
and the corresponding uncentered influence function has the form
$$\phi(O)
=
h(O) + \alpha(O)\{Y - f(O)\},$$
with 
$$f(O) = m(A,L),
\qquad
h(O) = \frac{I(V=1)}{P(V=1)} \frac{\partial}{\partial a} m(A,L).$$

```{r}
library(RieszCML)

set.seed(21)

n <- 5000

beta0 <- 1.0
beta1 <- 2.5
beta2 <- 1.2
sigma <- 0.5

L <- rnorm(n)
A <- rnorm(n, mean = L, sd = 1)
V <- as.integer(L > 0)
Y <- beta0 + beta1 * A + beta2 * L + rnorm(n, sd = sigma)

df <- data.frame(L = L, A = A, V = V, Y = Y)

truth_subgroup_ade <- beta1
truth_subgroup_ade

rc_subgroup_ade <- RieszCurve$new(
  nuis = function(data) {
    outcome_fit <- lm(Y ~ A + L, data = data)
    exposure_fit <- lm(A ~ L, data = data)

    m_hat <- predict(outcome_fit, newdata = data)

    beta1_hat <- unname(coef(outcome_fit)[["A"]])
    dm_da_hat <- rep(beta1_hat, nrow(data))

    muA_hat <- predict(exposure_fit, newdata = data)
    sigma2_hat <- summary(exposure_fit)$sigma^2

    # score_a = d/da log p(A|L) for Gaussian A|L
    score_a_hat <- -(data$A - muA_hat) / sigma2_hat

    pv_hat <- mean(data$V == 1)

    list(
      m = m_hat,
      dm_da = dm_da_hat,
      score_a = score_a_hat,
      pv = pv_hat
    )
  },
  alpha = ~ I(V == 1) / pv * score_a,
  f = ~ m,
  h = ~ I(V == 1) / pv * dm_da,
  ic_expr = ~ h + alpha * (Y - f)
)

fit_subgroup_ade <- riesz_estimate(df, rc_subgroup_ade)
fit_subgroup_ade

c(
  estimate = fit_subgroup_ade$estimate,
  truth = truth_subgroup_ade
)
```

## Variable Derivative Example

```{r}
set.seed(22)

n <- 5000

beta0 <- 1.0
beta1 <- 1.5
beta2 <- 0.8
beta3 <- 1.2
sigma <- 0.5

L <- rnorm(n)
A <- rnorm(n, mean = L, sd = 1)
V <- as.integer(L > 0)
Y <- beta0 + (beta1 + beta3 * L) * A + beta2 * L + rnorm(n, sd = sigma)

df2 <- data.frame(L = L, A = A, V = V, Y = Y)

truth_subgroup_ade_hetero <- beta1 + beta3 * sqrt(2 / pi)
truth_subgroup_ade_hetero

rc_subgroup_ade_hetero <- RieszCurve$new(
  nuis = function(data) {
    outcome_fit <- lm(Y ~ A + L + A:L, data = data)
    exposure_fit <- lm(A ~ L, data = data)

    m_hat <- predict(outcome_fit, newdata = data)

    betaA_hat <- unname(coef(outcome_fit)[["A"]])
    betaAL_hat <- unname(coef(outcome_fit)[["A:L"]])
    dm_da_hat <- betaA_hat + betaAL_hat * data$L

    muA_hat <- predict(exposure_fit, newdata = data)
    sigma2_hat <- summary(exposure_fit)$sigma^2
    score_a_hat <- -(data$A - muA_hat) / sigma2_hat

    pv_hat <- mean(data$V == 1)

    list(
      m = m_hat,
      dm_da = dm_da_hat,
      score_a = score_a_hat,
      pv = pv_hat
    )
  },
  alpha = ~ I(V == 1) / pv * score_a,
  f = ~ m,
  h = ~ I(V == 1) / pv * dm_da,
  ic_expr = ~ h + alpha * (Y - f)
)

fit_subgroup_ade_hetero <- riesz_estimate(df2, rc_subgroup_ade_hetero)
fit_subgroup_ade_hetero

c(
  estimate = fit_subgroup_ade_hetero$estimate,
  truth = truth_subgroup_ade_hetero
)
```


## Numerically Estimated Derivative


```{r}
# ------------------------------------------------------------
# Subgroup ADE with numerical approximation of dm/da
# ------------------------------------------------------------

library(RieszCML)

set.seed(22)

n <- 5000

# Data-generating mechanism
L <- rnorm(n)
A <- rnorm(n, mean = 0.5 * L, sd = 1)
V <- as.integer(L > 0)

# Nonlinear outcome regression in A
# m(a, l) = 1 + sin(a) + 0.5 * a * l + 0.3 * l^2
Y <- 1 + sin(A) + 0.5 * A * L + 0.3 * L^2 + rnorm(n, sd = 0.3)

df <- data.frame(L = L, A = A, V = V, Y = Y)

# ------------------------------------------------------------
# Truth
# ------------------------------------------------------------
# Here,
# d/da m(a,l) = cos(a) + 0.5 * l
#
# So the subgroup ADE is
# E[ cos(A) + 0.5 * L | V = 1 ]
#
# Since A | L ~ N(0.5 L, 1), there is no simple closed form for
# E[cos(A) | L > 0], so for the "truth" we approximate with a very large
# Monte Carlo sample from the known DGP.

truth_mc_n <- 2e6
L_big <- rnorm(truth_mc_n)
A_big <- rnorm(truth_mc_n, mean = 0.5 * L_big, sd = 1)
V_big <- as.integer(L_big > 0)

truth_subgroup_ade <- mean((cos(A_big) + 0.5 * L_big)[V_big == 1])
truth_subgroup_ade

nuis_subgroup_ade_numeric <- function(data, delta = 1e-4) {
  # Flexible working model for E[Y | A, L]
  # Here I use a simple spline basis in A and L to mimic a realistic workflow.
  outcome_fit <- lm(
    Y ~ splines::bs(A, df = 6) + splines::bs(L, df = 5) + A:L,
    data = data
  )

  # Predict m(A, L)
  m_hat <- predict(outcome_fit, newdata = data)

  # Numerically approximate dm/da using centered finite differences
  data_plus <- data
  data_minus <- data
  data_plus$A <- data_plus$A + delta
  data_minus$A <- data_minus$A - delta

  m_plus <- predict(outcome_fit, newdata = data_plus)
  m_minus <- predict(outcome_fit, newdata = data_minus)

  dm_da_hat <- (m_plus - m_minus) / (2 * delta)

  # Exposure score term for Gaussian A | L
  exposure_fit <- lm(A ~ L, data = data)
  muA_hat <- predict(exposure_fit, newdata = data)
  sigma2_hat <- summary(exposure_fit)$sigma^2

  # score_a = d/da log p(A | L)
  score_a_hat <- -(data$A - muA_hat) / sigma2_hat

  pv_hat <- mean(data$V == 1)

  list(
    m = m_hat,
    dm_da = dm_da_hat,
    score_a = score_a_hat,
    pv = pv_hat
  )
}

rc_subgroup_ade_numeric <- RieszCurve$new(
  nuis = nuis_subgroup_ade_numeric,
  alpha = ~ I(V == 1) / pv * score_a,
  f = ~ m,
  h = ~ I(V == 1) / pv * dm_da,
  ic_expr = ~ h + alpha * (Y - f)
)

fit_subgroup_ade_numeric <- riesz_estimate(df, rc_subgroup_ade_numeric)
fit_subgroup_ade_numeric

c(
  estimate = fit_subgroup_ade_numeric$estimate,
  truth = truth_subgroup_ade
)
```
